The Precambrian paleogeography of Laurentia

### 2020-11-05 Manuscript/Laurentia_Paleogeo_Manuscript.tex
edits prior to submission of revisions

As for the suturing of the Wyoming province to Laurentia (Fig. \ref{fig:Laurentia_map}), many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to the Hearne and Superior provinces while still undergoing continued translation up to ca. 1.80 Ga \citep{Whitmeyer2007a}. A contrasting view has been been proposed that the Wyoming and Medicine Hat provinces were not conjoined with the other Laurentia provinces until ca. 1.72 Ga \citep{Kilian2016b}. This interpretation is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca. 1.75 to 1.72 Ga \citep{Condit2015a} and ca. 1.72 tectonomagmatic activity in the Black Hills region \citep{Redden1990a}. However, the evidence for earlier orogenesis ca. 1.78 to 1.75 in the Black Hills \citep{Dahl1999a, Hrncir2017a}, as well as high-grade metamorphism as early as ca. 1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca. 1.72 Ga activity is a minor overprint on ca. 1.75 terminal suturing between the Wyoming and Superior provinces. Regardless, in both of these interpretations Wyoming is a later addition to Laurentia with final suturing post-dating ca. 1.82 Ga amalgamation of Archean provinces with the Trans-Hudson orogen further to the northeast. The Mojave province of southwestern Laurentia (Fig. \ref{fig:Laurentia_map}) comprises Paleoproterozoic gneiss that is interpreted based on isotopic data to include reworked Archean lithologies \citep{Bennett1987a}. more enigmatic both in terms of whether it constituents a province of Archean lithosphere or whether it should

@@

Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia. This change marks a shift in the predominant style of Laurentia's growth as subsequent crustal growth occurred dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). Determining the extent of these belts is complicated by poor exposure of them in the midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield. Major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca. 1.71 to 1.68 Ga Yavapai orogeny (Fig. \ref{fig:tectonic_history}). Yavapai orogenesis is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Karlstrom2001a}. Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}. These intrusions are hypothesized to have stabilized the juvenile accreted terranes that subsequently remained part of Laurentia \citep{Whitmeyer2007a}. Subsequent accretionary orogenesis of the ca. 1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism led to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}. Yavapai to Mazatzal-age accretionary orogenesis extended from the southwestern USA to northeast Canada in Labrador where it is called the Labradorian orogeny that is interpreted to have been active from ca. 1.71 to 1.60 Ga (Fig. \ref{fig:Laurentia_map}; \citealp{Gower1992a, Gower2008b}).

@@

In the central and southern Appalachians there are inliers of rocks that were metamorphosed during the Ottawan phase of the Grenvillian orogeny \citep{McLelland2013a}. On the basis of whole-rock Pb-isotope data, \cite{Loewy2003a} and \cite{Fisher2010a} proposed that these inliers are fragments of lithosphere of another continent that were transferred to Laurentia during the orogeny and left behind when the Iapetus Ocean formed. In particular, \cite{Fisher2010a} suggest that the Suns\'as orogen of Amazonia is the best match for southern and central Appalachian inliers. This positioning is consistent with a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly (Fig. \ref{fig:Grenville_reconstructions}; \citealp{Hoffman1991a, Evans2013b, Cawood2017a}). While the lack of ca. 1100 to 1000 Ma poles from Amazonia precludes a robust paleomagnetic test, this scenario is consistent with the available late Mesoproterozoic poles from Amazonia (ca. 1200 Nova Floresta pole and ca. 1150 Fortuna Formation pole; \citealp{DAgrella-Filho2021a}) as shown in \cite{Evans2013b}. In this paleogeographic scenario, the basement inliers of the Appalachian Orogen in the Blue Ridge region are interpreted to be the leading edge of Amazonia with initial collision ca. 1080 Ma initiating the Ottawan phase of the Grenvillian orogeny (Fig. \ref{fig:tectonic_history}). Subsequent separation of Amazonia would have lead to the formation of the Iapetus Ocean as Rodinia rifted apart. Departure of Amazonia potentially occurred as early as ca. 700 Ma in the Paleo-Iapetus Ocean model of  \citet{Robert2020a} in conjunction with rifting in eastern Laurentia. A significantly later separation ca. 600 Ma would be predicted if Amazonia were further north (present-day coordinates) along Laurentia's margin given the lack of evidence of riding until after  ca. 620 Ma north of the New York promontory.

@@

Southern Siberia and northern Laurentia have been proposed to be connected on the basis of a similar history of Paleoproterozoic collision \citep{Rainbird1998a}, matches in the age of mafic intrusives rocks interpreted as shared large igneous provinces from the Paleoproterozoic to the Neoporterozoic \cite{Ernst2016a}, and comparisons of paleomagnetic poles \citep{Evans2011a, Evans2016b}. U-Pb dates from mafic intrusive rocks are interpreted by \cite{Ernst2016a} as resulting from shared large igneous province  between southern Siberia and northern Laurentia from the time of Laurentia's amalgamation ca. 1.8 Ga all the way up to the time of the ca. 720 Ma Franklin large igneous province (LIP). Mafic intrusions and lavas that are grouped as the Irkutsk LIP in Siberia have been dated in Siberia to be similar in age to dates developed from the Franklin LIP \citep{Denyszyn2009a, Ernst2016a}. Comparisons of paleomagnetic poles between Laurentia and Siberia support such a tight and internally stable fit between southern Siberia and northern Laurentia from ca. 1.64 to at least ca. 0.76 Ga  \citep{Evans2016b}. Overlap in Laurentia and Siberia pole positions with such a reconstruction are achieved in the Statherian Period of the Paleoproterozoic (the ca. 1.64 Ga Nersa complex of Siberia compared to the Melville Bugt dikes of Laurentia), the Calymmian Period of the Mesoproterozoic (the ca. 1.50 Ga Anabar intrusions with the ca. 1.48 Ga St. Francois Mountains igneous province of Laurentia), the Stentian Period of the Mesoproterozoic (a number of roughly chronostratigraphically constrained Siberia poles with the 1.11 to 1.08 Ga Keweenawan Track of Laurentia) and the Tonian Period of the Neoproterozoic (the ca.  Kitoi pole of Siberia with Tonian Laurentia poles).  The correlation of the Malgina pole of Siberia with the Keweenawan Track gains additional support in that it correlates the normal-polarity Maya superchron \citep{Gallet2012a} with the Portage Lake normal-polarity zone \citep{Swanson-Hysell2019a} that is interpreted as a normal-polarity superchron (termed the Keweenawan Normal Superchron in \citealp{Driscoll2016a}). This correlation both works with the tight fit and resolves hemispheric ambiguity to put both cratons together in the northern hemisphere ca. 1100 Ma (Fig. \ref{fig:Grenville_reconstructions}).

Connections may have existed prior to Laurentia assembly such as the hypothesized connection between the Slave Province and the Tungus Province of Siberia which has the Thelon orogen correlated to Paleoproterozoic orogenesis in the Akitkan fold belt \citep{Condie1994a, Rainbird1998a, Evans2011a}. The emplacement of the Franklin LIP ca. 720 Ma could have initiated break-up of a joint Laurentia-Siberia continent ending the long Proterozoic connection between the cratons.

@@

Even without considering other continents, there is strong evidence both in Laurentia's geological and paleomagnetic record for differential plate tectonic motion between 2.2 and 1.8 Ga. The continued history of accretionary orogenesis and the evaluation of Laurentia's pole path in comparison to other continents from 1.8 Ga onward supports the continual operation of plate tectonics throughout the rest of the Proterozoic and Phanerozoic as well. While this evidence fits with the majority of interpretations of the timing of initiation of modern-style plate tectonics (see summary in \citealp{Korenaga2013a}), there continue to be arguments proposing that a stagnant lid persisted through the Mesoproterozoic Era (1.6 to 1.0 Ga) and into the Neoproterozoic  with plate tectonics not initiating until ca. 0.8 Ga \citep{Hamilton2011a, Stern2018a}. These arguments rest largely on the relative lack of Proterozoic low-temperature high-pressure metamorphic rocks such as blueschists that form in subduction zones \citep{Stern2013a}. An alternative interpretation for this lack of blueschists in the Proterozoic is that such a shift in metamorphic regime is the predicted result of secular evolution of mantle chemistry rather than a harbinger of the onset of plate tectonics \citep{Palin2015a}. While this line of evidence is intriguing, to argue that there was not differential plate tectonic motion in the Paleoproterozoic and Mesoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data. From a paleomagnetic perspective, there is strong support for independent and differential motion of the Slave and Superior provinces as is illustrated in Figure \ref{fig:Superior_Slave_recons}. From a geological perspective, the Trans-Hudson orogenic interval, the Grenville orogenic interval, and the Appalachian orogenic interval are all well-explained with a mobilistic interpretation that includes phases of accretionary followed by collisional orogenesis (Fig. \ref{fig:tectonic_history}). One could counter that this perspective results from a plate-tectonic-centric viewpoint that lacks creativity to see the record as resulting from other processes than modern-style plate tectonics. However, in addition to the broad geological record showing an amalgamation of terranes as would be expected to arise through plate tectonics, there is also an obducted ophiolite as well as eclogites preserved in the Trans-Hudson orogen \citep{Weller2017a}. These eclogites preserve evidence for high-pressure/low-temperature metamorphic conditions ca. 1.8 Ga. Similar to the Himalayan orogen, these rocks are interpreted to be the result of deep continental subduction and exhumation associated with convergent plate tectonics \citep{Weller2017a}. Outside of Laurentia, there are examples of Paleoproterozoic ecologites with geochemical affinity to oceanic crust such as that documented in the ca. 1.9 Ga Ubendian Belt of the Congo craton \citep{Boniface2012a}.

Another perspective on Proterozoic tectonics, is that the record is one of intermittent subduction \citep{Silver2008a, ONeill2013a}. In such a model, there are extended intervals with a stagnant lid alternating with intervals of differential plate motion. In particular, it has been argued that the Mesoproterozoic Era (1.6 to 1.0 Ga) is an interval when Earth was in a stagnant regime without mobile plate tectonics \citep{Silver2008a, ONeill2013a}. The long-lived accretionary history of Laurentia following the amalgamation of the Archean provinces is difficult to reconcile with such an interpretation (Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). The record of paleomagnetic poles also show that there was progressive motion of Laurentia through the Proterozoic (Figs. \ref{fig:Laurentia_paleolatitude} and \ref{fig:Laurentia_reconstructions}). Using data from Laurentia alone, however, it is difficult to ascertain whether this motion is due to plate tectonic motion or rotation of the entire solid Earth through true polar wander. True polar wander can lead to changing position relative to the spin axis even with a stagnant lid. One interval when the Laurentian paleomagnetic record demands that some of the motion is through differential plate tectonics is in the latest Mesoproterozoic. At that time, the pole path is very well-resolved with many high-quality paleomagnetic poles between 1110 and 1070 Ma (Table 2; \ref{fig:Laurentia_poles}). The progression of the poles requires rotation about an Euler pole that is distinct from a great circle path which would result if the motion were solely due to true polar wander \citep{Swanson-Hysell2019a}. These poles constrain rapid motion of Laurentia leading up to collisional orogenesis associated with the Grenvillian orogeny, as illustrated in Figure \ref{fig:Grenville_reconstructions}. These data provide strong evidence for differential plate motion at the time and are inconsistent with a stagnant lid. Rather, the orogenic interval of the Mesoproterozoic bears similarity with that of the Paleozoic and reveals Laurentia to have been a central player in the building of amalgamated continents associated with Rodinia and Pangea.

An additional constraint supporting ongoing plate tectonics throughout the Proterozoic comes from the paleomagnetic record --- in particular the paleomagnetic poles supported with baked contact tests (Fig. \ref{fig:baked_contact}). In a prolonged stagnant lid regime, there would not be sufficient heat flow across the core-mantle boundary to sustain a geodynamo \citep{Nimmo2000a, Buffett2000b}. One way to get insight into the ancient geomagnetic field is through paleointensity experiments on igneous rocks that enable estimates of ancient field strength to be developed.  Paleointensity data developed from units that are also included within the Laurentia paleomagnetic poles database indicate a significant geomagnetic field in the Neoarchean \citep{Selkin2000a} through to the Mesoproterozoic \citep{Macouin2006a, Sprain2018a}. However, paleointensity experiments are challenging and prone to failure due to alteration during laboratory or non-ideal rock magnetic behavior. As a result, it is significantly more challenging to develop reliable paleointensity data than the reliable paleodirectional data used to calculate paleomagnetic poles and the paleointensity database is sparser the compilation of reliable paleomagnetic poles. Paleodirectional data themselves can given insight into the presence of a significant geomagnetic field --- particularly those with positive baked contact tests. Baked contact tests indicate that, at the time of dike emplacement, there was an appreciable field such that both the cooling magma and the heated country rock in the vicinity of a dike were able to acquire a primary coherent magnetization direction. Additionally, since paleomagnetic poles are developed from many individual cooling units across a region, the similarity of the directions across an igneous province indicates that the magnetizations were dominantly acquired from the geomagnetic field rather than being influenced by local variable crustal magnetizations.  Therefore, the record of abundant positive baked contact tests (Fig. \ref{fig:baked_contact}) and coherent paleomagnetic poles (Table 2) supports the persistence of a geomagnetic field through the Paleoproterozoic and Mesoproterozoic which implies active plate tectonics that enabled sufficient core-mantle boundary heat flow to power the geodynamo.

@@

The paleogeographic record of Laurentia is rich in constraints through the Precambrian both in terms of the geological and geochronological constraints on tectonism and the record of paleomagnetic poles. Data from the Slave and Superior provinces of Laurentia provide what is arguably the strongest evidence of differential plate tectonics in the Rhyacian and Orosirian Periods of the Paleoproterozoic Era (2.3 to 1.8 Ga) leading up to the collision of these terranes during the Trans-Hudson orogeny. The collisions of these and other Archean provinces led to the formation of the core of Laurentia. Subsequent crustal growth occurred through multiple intervals of accretionary orogenesis through the late Paleoproterozoic and Mesoproterozoic until the continent-continent collision of the Grenvillian orogeny that was ongoing at the Mesoproterozoic-Neoproterozoic boundary (1.0 Ga). The lead-up to this orogeny was associated with rapid plate motion of Laurentia from high latitudes towards the equator recorded by the Logan Loop and Keweenawan Track of paleomagnetic poles. Following, a return to high latitudes as constrained by paleomagnetic poles of the Grenville Loop, Laurentia straddled the equator at the time of Cryogenian Snowball Earth glaciation as part of the Rodinia supercontinent. Rifting and passive margin development then isolated Laurentia in the early Paleozoic Era. Subsequent accretionary and collisional orogenesis occurred associated with the Appalachian orogenic interval with Laurentia first colliding with Avalonia-Baltica to become Laurussia and Laurussia then uniting with Gondwana to form Pangea. While the details of the conjugate continents are better reconstructed for this last Wilson cycle, the broad features of the Trans-Hudson, Grenvillian and Appalachian orogenic intervals bear similarities. In each case, accretionary collision of arc terranes was followed by continent-continent collision. The major difference is that the collisions of the Grenvillian and Appalachian orogenic intervals resulted in relatively minor crustal growth compared to the Trans-Hudson. Break-up following the Grenvillian and Appalachian orogenic intervals occurred along the same margin as collision while the major orogens of the Trans-Hudson orogenic interval have remained sutured. As a result, Laurentia has been a formidable continent for the past 1.8 billion years. As can be seen in the Chapters on Archean paleogeography \citep{Salminen2021b}, Nuna \citep{Elming2021a}, and Rodinia \citep{Evans2021b}, the constraints from Laurentia are at the center of paleogeographic models through the Precambrian and will continue to be as the next generation of paleogeographic models are developed.

@@

@@

@@

>>>>

As for the suturing of the Wyoming province to Laurentia (Fig. \ref{fig:Laurentia_map}), many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to the Hearne and Superior provinces while still undergoing continued translation up to ca. 1.80 Ga \citep{Whitmeyer2007a}. A contrasting view has been been proposed that the Wyoming and Medicine Hat provinces were not conjoined with the other Laurentia provinces until ca. 1.72 Ga \citep{Kilian2016b}. This interpretation is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca. 1.75 to 1.72 Ga \citep{Condit2015a} and ca. 1.72 tectonomagmatic activity in the Black Hills region \citep{Redden1990a}. However, the evidence for earlier orogenesis ca. 1.78 to 1.75 in the Black Hills \citep{Dahl1999a, Hrncir2017a}, as well as high-grade metamorphism as early as ca. 1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca. 1.72 Ga activity is a minor overprint on ca. 1.75 terminal suturing between the Wyoming and Superior provinces. Regardless, in both of these interpretations Wyoming is a later addition to Laurentia with final suturing post-dating ca. 1.82 Ga amalgamation of Archean provinces with the Trans-Hudson orogen further to the northeast.

@@

Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia. This change marks a shift in the predominant style of Laurentia's growth as subsequent crustal growth occurred dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). Determining the extent of these belts is complicated by poor exposure of them in the midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield. Major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca. 1.71 to 1.68 Ga Yavapai orogeny (Fig. \ref{fig:tectonic_history}). Yavapai orogenesis is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Karlstrom2001a}. Potentially associated with the Yavapai orogeny is the accretion of the Mojave province of southwestern Laurentia (Fig. \ref{fig:Laurentia_map}). The Mojave province comprises Paleoproterozoic gneiss that is interpreted based on isotopic data to include reworked Archean lithologies \citep{Bennett1987a}. It is unclear whether the Mojave province should be considered to be a distinct province of Archean lithosphere akin to other Archean provinces or rather it should dominantly be considered to be Yavapai arc terrane built upon minor fragments of Archean lithosphere \citep{Whitmeyer2007a}. Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}. These intrusions are hypothesized to have stabilized the juvenile accreted terranes that subsequently remained part of Laurentia \citep{Whitmeyer2007a}. Subsequent accretionary orogenesis of the ca. 1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism led to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}. Yavapai to Mazatzal-age accretionary orogenesis extended from the southwestern USA to northeast Canada in Labrador where it is called the Labradorian orogeny that is interpreted to have been active from ca. 1.71 to 1.60 Ga (Fig. \ref{fig:Laurentia_map}; \citealp{Gower1992a, Gower2008b}).

@@

In the central and southern Appalachians there are inliers of rocks that were metamorphosed during the Ottawan phase of the Grenvillian orogeny \citep{McLelland2013a}. On the basis of whole-rock Pb-isotope data, \cite{Loewy2003a} and \cite{Fisher2010a} proposed that these inliers are fragments of lithosphere of another continent that were transferred to Laurentia during the orogeny and left behind when the Iapetus Ocean formed. In particular, \cite{Fisher2010a} suggest that the Suns\'as orogen of Amazonia is the best match for southern and central Appalachian inliers. This positioning is consistent with a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly (Fig. \ref{fig:Grenville_reconstructions}; \citealp{Hoffman1991a, Evans2013b, Cawood2017a}). While the lack of ca. 1100 to 1000 Ma poles from Amazonia precludes a robust paleomagnetic test, this scenario is consistent with the available late Mesoproterozoic poles from Amazonia (ca. 1200 Nova Floresta pole and ca. 1150 Fortuna Formation pole; \citealp{DAgrella-Filho2021a}) as shown in \cite{Evans2013b}. In this paleogeographic scenario, the basement inliers of the Appalachian Orogen in the Blue Ridge region are interpreted to be the leading edge of Amazonia with initial collision ca. 1080 Ma initiating the Ottawan phase of the Grenvillian orogeny (Fig. \ref{fig:tectonic_history}). Subsequent separation of Amazonia would have lead to the formation of the Iapetus Ocean as Rodinia rifted apart. Departure of Amazonia potentially occurred as early as ca. 700 Ma in the Paleo-Iapetus Ocean model of  \citet{Robert2020a} in conjunction with rifting in eastern Laurentia. A significantly later separation ca. 600 Ma would be predicted if Amazonia were further north (present-day coordinates) along Laurentia's margin given the lack of evidence of rifting until after ca. 620 Ma north of the New York promontory.

@@

Southern Siberia and northern Laurentia have been proposed to be connected on the basis of a similar history of Paleoproterozoic collision \citep{Rainbird1998a}, matches in the age of mafic intrusives rocks interpreted as shared large igneous provinces from the Paleoproterozoic to the Neoproterozoic \cite{Ernst2016a}, and comparisons of paleomagnetic poles \citep{Evans2011a, Evans2016b}. U-Pb dates from mafic intrusive rocks are interpreted by \cite{Ernst2016a} as resulting from shared large igneous provinces between southern Siberia and northern Laurentia from the time of Laurentia's amalgamation ca. 1.8 Ga all the way up to the time of the ca. 720 Ma Franklin large igneous province (LIP). For example, mafic intrusions and lavas that are grouped as the Irkutsk LIP in Siberia have been dated in Siberia to be similar in age to dates developed from the Franklin LIP \citep{Denyszyn2009a, Ernst2016a}. Comparisons of paleomagnetic poles between Laurentia and Siberia support this tight and internally stable fit between southern Siberia and northern Laurentia from ca. 1.64 to at least ca. 0.76 Ga  \citep{Evans2016b}. Overlap in Laurentia and Siberia pole positions with such a reconstruction are achieved in the Statherian Period of the Paleoproterozoic (the ca. 1.64 Ga Nersa complex of Siberia compared to the Melville Bugt dikes of Laurentia), the Calymmian Period of the Mesoproterozoic (the ca. 1.50 Ga Anabar intrusions with the ca. 1.48 Ga St. Francois Mountains igneous province of Laurentia), the Stentian Period of the Mesoproterozoic (a number of roughly chronostratigraphically constrained Siberia poles with the 1.11 to 1.08 Ga Keweenawan Track of Laurentia) and the Tonian Period of the Neoproterozoic (the ca.  Kitoi pole of Siberia with Tonian Laurentia poles).  The correlation of the Malgina pole of Siberia with the Keweenawan Track gains additional support in that it correlates the normal-polarity Maya superchron \citep{Gallet2012a} with the Portage Lake normal-polarity zone \citep{Swanson-Hysell2019a} that is interpreted as a normal-polarity superchron (termed the Keweenawan Normal Superchron in \citealp{Driscoll2016a}). This correlation both works with the tight fit and resolves hemispheric ambiguity to put both cratons together in the northern hemisphere ca. 1100 Ma (Fig. \ref{fig:Grenville_reconstructions}).

Connections between Archean provinces may have also existed prior to Laurentia assembly such as the hypothesized connection between the Slave Province and the Tungus Province of Siberia which has the Thelon orogen correlated to Paleoproterozoic orogenesis in the Akitkan fold belt \citep{Condie1994a, Rainbird1998a, Evans2011a}. The emplacement of the Franklin LIP ca. 720 Ma could have initiated break-up of a joint Laurentia-Siberia continent ending the long Proterozoic connection between the cratons.

@@

Even without considering other continents, there is strong evidence both in Laurentia's geological and paleomagnetic record for differential plate tectonic motion between 2.2 and 1.8 Ga. The continued history of accretionary orogenesis and the evaluation of Laurentia's pole path in comparison to other continents from 1.8 Ga onward supports the continual operation of plate tectonics throughout the rest of the Proterozoic and Phanerozoic as well. While this evidence fits with the majority of interpretations of the timing of initiation of modern-style plate tectonics (see summary in \citealp{Korenaga2013a}), there continue to be arguments proposing that a stagnant lid persisted through the Mesoproterozoic Era (1.6 to 1.0 Ga) and into the Neoproterozoic  with plate tectonics not initiating until ca. 0.8 Ga \citep{Hamilton2011a, Stern2018a}. These arguments rest largely on the relative lack of Proterozoic low-temperature high-pressure metamorphic rocks such as blueschists that form in subduction zones \citep{Stern2013a}. An alternative interpretation for this lack of blueschists in the Proterozoic is that such a shift in metamorphic regime is the predicted result of secular evolution of mantle chemistry rather than a harbinger of the onset of plate tectonics \citep{Palin2015a}. While this line of evidence is intriguing, to argue that there was not differential plate tectonic motion in the Paleoproterozoic and Mesoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data. From a paleomagnetic perspective, there is strong support for independent and differential motion of the Slave and Superior provinces as is illustrated in Figure \ref{fig:Superior_Slave_recons}. From a geological perspective, the Trans-Hudson orogenic interval, the Grenville orogenic interval, and the Appalachian orogenic interval are all well-explained with a mobilistic interpretation that includes phases of accretionary orogenesis followed by collisional orogenesis (Fig. \ref{fig:tectonic_history}). One could counter that this perspective results from a plate-tectonic-centric viewpoint that lacks creativity to see the record as resulting from other processes than modern-style plate tectonics. However, in addition to the broad geological record showing an amalgamation of terranes as would be expected to arise through plate tectonics, there are also Proterozoic obducted ophiolites that are well-explained through accretionary plate tectonicsas as well as eclogites such as those preserved in the Trans-Hudson orogen \citep{Weller2017a}. These eclogites preserve evidence for high-pressure/low-temperature metamorphic conditions ca. 1.8 Ga. Similar to the Himalayan orogen, these rocks are interpreted to be the result of deep continental subduction and exhumation associated with convergent plate tectonics \citep{Weller2017a}. Outside of Laurentia, there are examples of Paleoproterozoic ecologites with geochemical affinity to oceanic crust such as that documented in the ca. 1.9 Ga Ubendian Belt of the Congo craton \citep{Boniface2012a}. Mesoproterozoic ophiolites were also obducted to Laurentia such as the Pyrites ophiolite complex within the Shawinigan orogen and the Coal Creek Domain of the Llano uplift \citep{Chiarenzelli2011a, McLelland2013a}.

Another perspective on Proterozoic tectonics, is that the record is one of intermittent subduction \citep{Silver2008a, ONeill2013a}. In such a model, there are extended intervals with a stagnant lid alternating with intervals of differential plate motion. In particular, it has been argued that the Mesoproterozoic Era (1.6 to 1.0 Ga) is an interval when Earth was in a stagnant regime without mobile plate tectonics \citep{Silver2008a, ONeill2013a}. The long-lived accretionary history of Laurentia following the amalgamation of the Archean provinces is difficult to reconcile with such an interpretation (Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). The record of paleomagnetic poles also show that there was progressive motion of Laurentia through the Proterozoic (Figs. \ref{fig:Laurentia_paleolatitude} and \ref{fig:Laurentia_reconstructions}). Using data from Laurentia alone, however, it is difficult to ascertain whether this motion is due to plate tectonic motion or rotation of the entire solid Earth through true polar wander. True polar wander can lead to changing position relative to the spin axis even with a stagnant lid. One interval when the Laurentian paleomagnetic record demands that some of the motion is through differential plate tectonics is in the latest Mesoproterozoic. At that time, the pole path is very well-resolved with many high-quality paleomagnetic poles between 1110 and 1070 Ma (Table 2; \ref{fig:Laurentia_poles}). The progression of the poles requires rotation about an Euler pole that is distinct from a great circle path which would result if the motion were solely due to true polar wander \citep{Swanson-Hysell2019a}. These poles constrain rapid motion of Laurentia leading up to collisional orogenesis associated with the Grenvillian orogeny, as illustrated in Figure \ref{fig:Grenville_reconstructions}. These data provide strong evidence for differential plate motion at the time and are inconsistent with a stagnant lid. Rather, the orogenic interval of the Mesoproterozoic (1.6 to 1.0 Ga) bears similarity with that of the Paleozoic (0.54 to 0.25 Ga) and reveals Laurentia to have been a central player in amalgamation of continents associated with both Rodinia and Pangea.

An additional constraint supporting ongoing plate tectonics throughout the Proterozoic comes from the paleomagnetic record --- in particular the paleomagnetic poles supported with baked contact tests (Fig. \ref{fig:baked_contact}). In a prolonged stagnant lid regime, there would not be sufficient heat flow across the core-mantle boundary to sustain a geodynamo \citep{Nimmo2000a, Buffett2000b}. One way to get insight into the ancient geomagnetic field is through paleointensity experiments on igneous rocks that enable estimates of ancient field strength to be developed.  Paleointensity data developed from units that are also included within the Laurentia paleomagnetic poles database indicate a significant geomagnetic field in the Neoarchean \citep{Selkin2000a} through to the Mesoproterozoic \citep{Macouin2006a, Sprain2018a}. However, paleointensity experiments are challenging and prone to failure due to alteration during laboratory hearing or non-ideal rock magnetic behavior. As a result, it is significantly more challenging to develop reliable paleointensity data than to develop the reliable paleodirectional data used to calculate paleomagnetic poles. Therefore, the paleointensity database is sparser the compilation of reliable paleomagnetic poles. Paleodirectional data themselves can given insight into the presence of a significant geomagnetic field --- particularly those with positive baked contact tests. Baked contact tests indicate that, at the time of dike emplacement, there was an appreciable field such that both the cooling magma and the heated country rock in the vicinity of a dike were able to acquire a primary coherent magnetization direction. Additionally, since paleomagnetic poles are typically developed from many individual cooling units across a region, the similarity of the directions across an igneous province indicates that the magnetizations were dominantly acquired from the geomagnetic field rather than being influenced by local variable crustal magnetizations.  Therefore, the record of abundant positive baked contact tests (Fig. \ref{fig:baked_contact}) and coherent paleomagnetic poles (Table 2) supports the persistence of a geomagnetic field through the Paleoproterozoic and Mesoproterozoic which implies active plate tectonics that enabled sufficient core-mantle boundary heat flow to power the geodynamo.

@@

The paleogeographic record of Laurentia is rich in constraints through the Precambrian both in terms of the geological and geochronological data on tectonism and the record of paleomagnetic poles. Data from the Slave and Superior provinces of Laurentia provide what is arguably the strongest evidence of differential plate tectonics in the Rhyacian and Orosirian Periods of the Paleoproterozoic Era (2.3 to 1.8 Ga) leading up to the collision of these terranes during the Trans-Hudson orogeny. The collisions of these and other Archean provinces led to the formation of the core of Laurentia. Subsequent crustal growth occurred through multiple intervals of accretionary orogenesis through the late Paleoproterozoic and Mesoproterozoic until the continent-continent collision of the Grenvillian orogeny that was ongoing at the Mesoproterozoic-Neoproterozoic boundary (1.0 Ga). The lead-up to this orogeny was associated with rapid plate motion of Laurentia from high latitudes towards the equator recorded by the Logan Loop and Keweenawan Track of paleomagnetic poles. Following a return to high latitudes, as constrained by paleomagnetic poles of the Grenville Loop, Laurentia straddled the equator at the time of Cryogenian Snowball Earth glaciation as part of the Rodinia supercontinent. Rifting and passive margin development then isolated Laurentia in late Ediacaran period and into the early Paleozoic Era. Subsequent accretionary and collisional orogenesis occurred associated with the Appalachian orogenic interval with Laurentia first colliding with Avalonia-Baltica to become Laurussia and Laurussia then uniting with Gondwana to form Pangea. While the details of the conjugate continents are better reconstructed for this last Wilson cycle, the broad features of the Trans-Hudson, Grenvillian and Appalachian orogenic intervals bear similarities. In each case, accretionary collision of arc terranes was followed by continent-continent collision. The major difference is that the collisions of the Grenvillian and Appalachian orogenic intervals resulted in relatively minor crustal growth compared to the Trans-Hudson orogeny. Break-up following the Grenvillian and Appalachian orogenic intervals occurred along the same margin as collision while the major orogens of the Trans-Hudson orogenic interval have remained sutured. As a result, Laurentia has been a formidable continent for the past 1.8 billion years. As can be seen in the Chapters on Archean paleogeography \citep{Salminen2021b}, Nuna \citep{Elming2021a}, and Rodinia \citep{Evans2021b}, the constraints from Laurentia are at the center of paleogeographic models through the Precambrian and will continue to be as the next generation of paleogeographic models are developed.

@@

\noindent\textbf{Cryogenian Period } The geologic period that lasted from ca. 717 to 635 million years ago during which time there were two global glaciations. The start of the period is provisionally defined as the first evidence of low-latitude glaciation. It is the second geologic period of the Neoproterozoic Era being preceded by the Tonian Period and followed by the Ediacaran Period.

\noindent\textbf{Ediacaran Period } The third geologic period of the Neoproterozoic Era from ca. 635 to 539 Ma million years ago. It is the final period of the Proterozoic Eon and is followed by the Cambrian Period.

@@

\noindent\textbf{ophiolite } Oceanic lithosphere that as been accreted onto continental lithosphere.

@@

\noindent\textbf{Tonian Period } The first geologic period of the Neoproterozoic Era from ca. 1000 to 717 Ma million years ago. It followed by the Cryogenian Period.
